Formal models of biological networks in ANIMO 2

### 2013-12-13 Paper.tex
Added more on model checking, started working on better conclusions

\begin{abstract} % abstract <= 350 words. Headers: background, results, conclusions
\parttitle{Background}
The large amount of data with which an experimental biologist
must deal makes it largely impossible to reason on
the dynamics of complex systems without relying on computational
support.
In order to be fully profitable, computational modelling tools
must be both accessible and powerful, providing a means to
formalise and analyse experimental outcomes without requiring
additional training.
\parttitle{Results}
We show that the software tool ANIMO (Analysis of Networks with Interactive MOdelling) can be used
in different modelling tasks, and illustrate how this can be an asset for the biological researcher.
ANIMO models for \emph{Drosophila melanogaster} circadian clock and signal
transduction events downstream of TNF$\alpha$ and EGF in HT-29 human
colon carcinoma cells are used as case studies to illustrate the
possibilities offered by the tool.
\parttitle{Conclusions}
The software tool ANIMO, recently updated to version 2, allows the biologist to develop
a formal model through a user friendly interface,
achieving useful insight without the need to study any mathematical formalism.
\end{abstract}

@@

The case study we use as a testbed is shown in Figure~\ref{fig:cytoscape},
and is a model signaling events downstream of growth factors EGF (epidermal growth factor)
and NGF (nerve growth factor) in PC12 cells.
The topology for this network was proposed in~\cite{egf-ngf}, and it was analyzed
with an ANIMO model, reproducing the experimentally 
observed ERK (extracellular signal-regulated kinase) activity changes~\cite{animo-bibe}.
In particular, a 10 minutes stimulation with EGF resulted in transient behavior (i.e. peak-shaped,
see also the graph in Fig.~\ref{fig:cytoscape}), while NGF stimulation led to sustained activity (see Fig.~\ref{fig:comparison-graph-animo}).

\subsubsection*{Simulation-based analysis}
The most basic analysis that can be performed with ANIMO is based on simulating the behaviour
of the network during a given period of time (see also the analyses
of the Drosphila and TNF-$\alpha$/EGF models above). As this type of analysis is also normally
used when looking for a match between model parameters and experimental data, the computation
speed of a simulation run deeply affects the level of interactivity perceived by the user.
In order to make
the modeling approach in ANIMO as smooth as possible, it is desirable to decrease dead times,
i.e. significantly reduce simulation costs.
% UPPAAL is used in ANIMO to obtain simulation runs from a \tas\ model. While that process is
% very fast, 
% In this experiment, we query UPPAAL for simulation runs on the models
Our first performance test compares the execution time for simulation runs on the \tas\ models
generated by ANIMO applying the old (i.e., ANIMO 1.0) and new (ANIMO 2.0) modeling approaches to the case study.
To define the initial state of the model, we consider the starting condition to be the treatment with 50~ng/ml NGF, which translates into
setting the initial activity level of nodes NGF to 15/15, and EGF to 0/15. We have chosen this configuration as the
model is expected to continue reacting indefinitely (sustained ERK activity), without reaching a deadlock state where all reactions are inactive.
This makes the model more demanding when generating the state space
because having more reactions occur means having more states in the transition system underlying the \tas\ model.
Table~\ref{tab:sim-100} illustrates the computation time and memory usage when performing 100 simulation runs on each of the
two considered models. Computing the simulation runs took about $92 \%$ less time with the new version of ANIMO, using $87 \%$
less memory. This decrease in computation time for long simulation runs brings the approach nearer to the idea
of interactive exploration of a network. This advantage is particularly noticeable when analysing
more complex networks.

@@

The technique of model checking consists in verifying the truth of some properties
that describe the behaviour of a model. This is usually done by exploring every
possible execution path of the model starting from its initial configuration. The set of all
the configurations that can be reached by a model during all its possible executions
is called \emph{state space}. The properties on the evolution of the model are
usually expressed in a formal language such as computation tree logic (CTL,~\cite{ctl}),
which allows to formally describe how and when a property should hold. Examples of such
properties can be expressed in natural language as: ``the activity of node A
stays always above 20'', ``it is possible to reach a state where nodes A and B are both
active for at least 60 \%'', ``whenever node A becomes active for at least 20 \%, node
B becomes active for at least 30 \%''.

@@

The analysis of query (1) returned false and all other queries returned true.
In particular, query (1) confirms that under EGF treatment no other activity is observed in the model after the initial peak,
while query (2) confirms that with NGF activity continues indefinitely.
Moreover, queries (3) - (6) confirm the result of the simulations shown in~\cite{animo-bibe}, with NGF treatment leading to sustained ERK activity.
The model checking performance of UPPAAL using the reaction- and reactant-centered model types is shown in Table~\ref{tab:model-checking}.

\subsubsection*{Analysis of the results}
Note that we focus most of our attention on the NGF treatment, as with this configuration the model
was hypothesized to keep reacting permanently (sustained ERK activity). The results from queries (1) and (2)
show that this is indeed the case, as EGF treatment leads to all reactions dying out after a number of steps (i.e., deadlock due to all
automata being in the {\sf not\_reacting} location).
This means that the queries (3) - (6) deal with a larger state space, leading to presumably lower model checking performances.
Requesting a full inspection of the state space
as we do when using a query of the type {\tt A[]$\phi$} returning true in cases (5) and (6), allows us to indirectly compare the state space size of the two
model versions. As the hundreds-fold computation time improvements in Table~\ref{tab:model-checking} show, the reactant-centered model produces
indeed a noticeably smaller state space, allowing for a higher level of interactivity also when performing non-trivial model checking.

Moreover, our experiments point out that the reactant-centered approach considerably lowers the memory
requirements for the model. This is not only due to the absence of possibly large precomputed time tables,
which can contain up to ten thousand elements each.\footnote{In the reaction-centered model, having a reaction depend on two 100-levels reactants
leads to the generation of two time tables containing $100 \times 100 = 10\,{}000$ elements each.}
Indeed, this point was further investigated by implementing a reaction-centered model which
avoids the use of tables and instead makes on-the-fly computations of the time bounds with the same number representation as in the reactant-centered model.
This resulted in improved performances in the cases of reachability and simulation-based queries, with memory requirements similar to the ones for
the reactant-centered model. However, in all other cases a much larger amount of memory (more than 2 Gb) was used
with respect to the table-based implementation of the same model, without leading to appreciable benefits in terms of execution time:
in some cases performances were noticeably deteriorated. These findings support the intuitive idea that a reactant-centered approach
leads in general to fewer events in a model, and to a smaller state space.

Note that while the network we consider for the case study is not particularly large, the gain in performances is still considerable
and promising in the perspective of analyzing more complex networks.

\subsection*{Model checking for bigger models}
When a model becomes particularly complex, and it represents many events occurring concurrently,
the state space of the model tends to grow exponentially in size, dramatically
increasing the time required for its analysis. To avoid this state space explosion,
\emph{statistical model checking} is commonly applied. This type of analysis is based on statistical
considerations and allows to obtain answers with a given confidence level in far less time
than most exact model checking techniques.

@@

Going beyond the user interface, there are a number of ``pros and cons'' for using ANIMO and \tas\ 
in the biological context. First and foremost, being \tas\ an executable formal language,
a state space can be derived from a \tas\ model. This means that state space-related analyses
such as model checking can be performed on \tas: as explained in the Results section, this can
be done directly in ANIMO, bypassing the need for a deeper understanding of \tas\ while
indirectly applying a powerful model checking tool such as UPPAAL.

The smaller amount of parameters needed to build an ANIMO model, if compared e.g. to
a differential equation-based approach, makes the modelling process faster, allowing
the biologist to obtain interesting results in less time. However, a less
parameter-intensive modelling approach is also less precise: in our case we abstract
sequences of biochemical reactions into activation or inactivation interactions,
simplifying any relation between two entities in a network as either activating or
inhibiting. While this allows for some simplifications in the networks and the underlying
formal models, it does not allow to represent in full detail more specific concepts
as reactant concentration or reaction affinity. In case such concepts are of
primary importance, we encourage the interested reader to turn to more precise
modelling approaches, such as those based on differential equations.
In particular, all reactants modelled in ANIMO are assumed to be present at the same
(unitary) concentration. While this assumption is not always applicable, it encourages
abstract thought: many biological processes can be represented as networks driven by
\emph{activity-based} interactions. Even if with a limited scope, ANIMO can be applied
also in the analysis of metabolic processes, considering the production of protein $A$ as
the activation of node $A$, with degratation being represented as inhibition. As stated before,
it cannot be expected from such models to be truly faithful to their respective biological
processes, but they can still be an useful tool.

\subsection*{Final remarks}
% The ultimate aim of research projects is to solve a problem or get
% the answer to a scientific question. In biology, an in-depth understanding of
% the relevant biological system is an important step towards this goal. Successive 
% repetitions of the empirical cycle result in a stepwise increase in understanding,
% until the goal is reached. For complex biological systems, computational modelling is 
% indispensable in this process. The mere act of creating 
% a computational model based on prior knowledge, experimental data and hypotheses 
% assists in gaining more insight in the system. 
% 
% %Furthermore, \emph{in silico} experiments 
% %with the model can be used to make predictions that cannot be made by the human brain. 
% 
% We developed a new modelling approach by proposing a series of abstractions from the detailed 
% molecular mechanisms of biological systems. These abstractions reduce the need for kinetic 
% parameters, while preserving enough expressivity for a useful description of the dynamic 
% behaviour of biological networks. A novel modelling tool, ANIMO, allows 
% effective use of this approach and enables an intuitive construction of formal models.

ANIMO is not the first modelling tool to provide an interface to a
modelling formalism. Such interfaces exist in many other tools (see Suppl. Tab.~\ref{tab:tool-comparison}). With its
focus on user-friendliness and intuitive modelling, ANIMO's main contribution lies 
in making computational modelling more accessible to experts in biology.
Making use of the visual
interface provided by Cytoscape, network representations subscribe to biological conventions. 
Model parameters are kept to a minimum and can be directly accessed by mouse-clicking on 
nodes and edges. Because of the automatic translation of the network topology and 
user-defined parameters into an underlying formal model, training in the use of formal methods 
is not needed. In Supplementary Section~\ref{suppl:comparison-table}, a more in-depth
comparison between ANIMO and other modeling tools is given. For this comparison we selected a tool
for each of the most commonly used formalisms, and used criteria with a strong focus on 
user-friendliness.

In Section~\ref{sec:animo-drosophila}, we described the construction of an ANIMO
model of the ciracadian clock in \emph{Drosophila Melanogaster}. This model
captured the dynamics of the regulatory network and led to similar 
conclusions as an ODE model that had been
published previously~\cite{drosophila-ode-model}. This finding supports the use of
the series of modelling abstractions that we proposed. The biggest
difference between the construction of these models is that the model by~\cite{drosophila-ode-model}
is constructed by writing a system of mathematical equations, together
with an algorithm for simulation. In ANIMO, instead, a number of network
nodes is drawn for the molecules involved. 
These nodes are then linked by directed
interactions that represent cause-and-effect relationships, with a single parameter 
that defines the strength of each
interaction. This is a more intuitive approach to construct a model.
Further contributing to an interactive modelling process
is the compositionality of the model. Each node in the network
can be disabled at any time by the user, or extra nodes can be added,
without having to change any of the existing interactions.

In Section~\ref{subsec:case-study-larger}, we showed the construction of an executable model
of signalling events downstream of
TNF$\alpha$ and EGF in human colon carcinoma cells. This data set has been used for
previous modelling studies, based on partial least-squares regression and fuzzy 
logic~\cite{pathway-leastsquare,pathway-fuzzy}.
The partial least-squares regression model describes an abstract data-driven model 
that uses statistical correlations
to relate signal transduction events to various cellular decisions. This type of modelling is
very useful in uncovering new and unexpected relations. It is also successful in making
predictions, but gives little direct in the dynamic behaviour of the network. Fuzzy
logic analysis led to a model that gave a better fit to the dynamic network behaviour than
discrete logic (Boolean) models. Inspection of the inputs to the logical gates that were used
to model protein behaviour led to the prediction of novel interactions between proteins,
showing the usefulness of this approach. For most of the proteins, such as JNK1, time was
used as an input parameter. For example, the logical gates ``if TNF$\alpha$ is high
\emph{AND} time is low, then JNK1 is high'' and ``if TNF$\alpha$ is high \emph{AND} time is
high, then JNK1 is low'' were used to
describe the dynamic behaviour of JNK1. Although this leads to a representative
description of the dynamic behaviour of JNK1, peaks in protein activity at early time points were
not reproduced by the fuzzy logic model. Moreover, it gives no insight in the molecular interactions 
that are involved in activation or inhibition.

@@

\section*{Conclusions}
We have presented here how the ANIMO tool 
was improved to provide an interactive approach to model checking.
Thanks to the increased performances of the new reactant-centered modeling approach,
we are able to obtain answers to model checking queries in a matter of seconds.
Coherently with the user friendliness of the tool, we ensure that the biologist can access the features
of model checking without the need to directly deal with \tas\ models.
In this way, ANIMO acts as an intermediary between the biologist and a formal
representation of biological signaling pathways, letting the experts concentrate
on investigating the mechanisms of cellular responses.

In order to enforce the concept of user interaction as a primary focus of the tool, we plan to extend
ANIMO by adding support for parameter sensitivity analysis and automatic parameter sweeps.
We already allow for only one numeric parameter per reaction, but an automatic way of adjusting such
parameters to fit experimental data will allow the user to concentrate more on defining a sensible topology for a pathway.
Such an automatic search could be limited by setting specific bounds on reaction rates based on previous knowledge,
or on a trade-off with performances.
Moreover, inspired by works on automata learning~\cite{test-based-modelling}, we plan to add also the possibility
to automatically derive a network topology based on experimental data and user-defined restrictions based on
previous knowledge.

>>>>

\begin{abstract} % abstract <= 350 words. Headers: background, results, conclusions
\parttitle{Background}
The large amount of data with which an experimental biologist
must deal makes it nearly impossible to reason on
the dynamics of complex systems without relying on computational
support.
In order to be fully profitable, computational modelling tools
must be both accessible and powerful, providing a means to
formalise and analyse experimental outcomes without requiring
additional training.
\parttitle{Results}
We show that the software tool ANIMO (Analysis of Networks with Interactive MOdelling) can be used
in different modelling tasks, and illustrate how this can be an asset for the biological researcher.
ANIMO models for \emph{Drosophila melanogaster} circadian clock and signal
transduction events downstream of TNF$\alpha$ and EGF in HT-29 human
colon carcinoma cells are used as case studies to illustrate the
possibilities offered by the tool.
\parttitle{Conclusions}
The software tool ANIMO, recently updated to version 2, allows the biologist to develop
a formal model through a user friendly interface,
achieving useful insight without the need to study any mathematical formalism.
\end{abstract}

@@

The case study we use as a testbed is shown in Figure~\ref{fig:cytoscape},
and is a model of signaling events downstream of growth factors EGF (epidermal growth factor)
and NGF (nerve growth factor) in PC12 cells.
The topology for this network was proposed in~\cite{egf-ngf}, and it was analyzed
with an ANIMO model, reproducing the experimentally 
observed ERK (extracellular signal-regulated kinase) activity changes~\cite{animo-bibe}.
In particular, a 10 minutes stimulation with EGF resulted in transient behavior (i.e. peak-shaped,
see also the graph in Fig.~\ref{fig:cytoscape}), while NGF stimulation led to sustained activity (see Fig.~\ref{fig:comparison-graph-animo}).

\subsubsection*{Simulation-based analysis}
The most basic analysis that can be performed with ANIMO is based on simulating the behaviour
of the network during a given period of time (see also the analyses
of the Drosphila and TNF-$\alpha$/EGF models above). As this type of analysis is also normally
used when looking for a match between model parameters and experimental data, the computation
speed of a simulation run deeply affects the level of interactivity perceived by the user.
In order to make
the modeling approach in ANIMO as smooth as possible, it is desirable to decrease dead times,
i.e. significantly reduce simulation costs.
% UPPAAL is used in ANIMO to obtain simulation runs from a \tas\ model. While that process is
% very fast, 
% In this experiment, we query UPPAAL for simulation runs on the models
Our first performance test compares the execution time for simulation runs on the \tas\ models
generated by ANIMO applying the old (i.e., ANIMO 1.0) and new (ANIMO 2.0) modeling approaches to the case study.
To define the initial state of the model, we consider the starting condition to be the treatment with 50~ng/ml NGF, which translates into
setting the initial activity level of nodes NGF to 15/15, and EGF to 0/15. We have chosen this configuration as the
model is expected to continue reacting indefinitely (sustained ERK activity), without reaching a deadlock state where all reactions are inactive.
This makes the model more demanding when generating the state space
because having more reactions occur means having more states in the transition system underlying the \tas\ model.
Table~\ref{tab:sim-100} illustrates the computation time and memory usage when performing 100 simulation runs on each of the
two considered models. Computing the simulation runs took about $92 \%$ less time with the new version of ANIMO, using $87 \%$
less memory. This decrease in computation time for long simulation runs brings the approach nearer to the idea
of interactive exploration of a network. This advantage is particularly noticeable when analysing
more complex networks.
{\Large [[Add also lines for drosophila and TNFa-EGF models, that are larger and make the performance increment look even more impressive]]}

@@

The technique of model checking consists in verifying the truth of some properties
that describe the behaviour of a model. This is usually done by exploring every
possible execution path of the model starting from its initial configuration. The set of all
the configurations that can be reached by a model during all its possible executions
is called \emph{state space}. The properties on the evolution of the model are
usually expressed in a formal language such as computation tree logic (CTL,~\cite{ctl}),
which allows to formally describe how and when a property should hold. Examples of such
properties can be expressed in natural language as: ``the activity of node A
stays always above 20'', ``it is possible to reach a state where the activity of both nodes A and B
is at least 60'', ``whenever the activity of node A surpasses 20, the activity of node
B will eventually reach at least 40''.

@@

The analysis of query (1) returned false and all other queries returned true.
In particular, query (1) confirms that under EGF treatment no other activity is observed in the model after the initial peak,
while query (2) confirms that with NGF activity continues indefinitely.
Moreover, queries (3) - (6) confirm the result of the simulations shown in~\cite{animo-bibe}, with NGF treatment leading to sustained ERK activity.
The model checking performance of UPPAAL comparing the two versions of ANIMO is shown in Table~\ref{tab:model-checking}.

\subsubsection*{Analysis of the results}
Note that we focus most of our attention on the NGF treatment, as with this configuration the model
was hypothesized to keep reacting permanently (sustained ERK activity). The results from queries (1) and (2)
show that this is indeed the case, as EGF treatment leads to all reactions dying out after a number of steps.
%(i.e., deadlock due to all automata being in the {\sf not\_reacting} location).
This means that the queries (3) - (6) deal with a larger state space, leading to presumably lower model checking performances.
Requesting a full inspection of the state space
as we do when using a query of the type {\tt A[]$\phi$} returning true in cases (5) and (6), allows us to indirectly compare the state space size of the two
model versions. As the hundreds-fold computation time improvements in Table~\ref{tab:model-checking} show, the new model produces
indeed a noticeably smaller state space, allowing for a higher level of interactivity also when performing non-trivial model checking.

Moreover, our experiments point out that ANIMO 2.0 considerably lowers the memory
requirements for the model. This is not only due to the absence of possibly large precomputed time tables,
which can contain up to ten thousand elements each.\footnote{In the model used in ANIMO 1.0, having a reaction depend on two 100-levels reactants
lead to the generation of two time tables containing $100 \times 100 = 10\,{}000$ elements each.}
% Indeed, this point was further investigated by implementing a reaction-centered model which
% avoids the use of tables and instead makes on-the-fly computations of the time bounds with the same number representation as in the reactant-centered model.
% This resulted in improved performances in the cases of reachability and simulation-based queries, with memory requirements similar to the ones for
% the reactant-centered model. However, in all other cases a much larger amount of memory (more than 2 Gb) was used
% with respect to the table-based implementation of the same model, without leading to appreciable benefits in terms of execution time:
% in some cases performances were noticeably deteriorated. These findings support the intuitive idea that a reactant-centered approach
% leads in general to fewer events in a model, and to a smaller state space.

Note that while the network we consider for the case study is not particularly large, the gain in performances is still considerable
and promising in the perspective of analyzing more complex networks.
{\Large [[Add some example queries for the TNFa-EGF model??]]}

\subsection*{Model checking for bigger models}
When a model becomes particularly complex, and it represents many events occurring concurrently,
the state space of the model tends to grow exponentially in size, dramatically
increasing the time required for its analysis. To avoid this \emph{state space explosion},
other approaches to model checking are necessary. One of such approaches is based on statistical
considerations and allows to obtain approximated answers with a given confidence level in far less time
than most exact model checking techniques: it is called statistical model checking. As the newer
versions of UPPAAL include also a statistical model checking engine~\cite{uppaal-smc}, we are
able to obtain answers in case a model displays stochastic behavior. Models of this type 
usually represent non-deterministic behavior with by defining a probabilistic distribution for the
choice of alternative evolution paths. In ANIMO, it is possible to use a uniform probabilistic choice
to decide the value of an interaction speed (chosen within the user-given lower and upper bounds)
and the initial activity level of a node. If no such variability is introduced, ANIMO models are deterministic,
i.e. they will always exhibit the same behavior. As an example application of the statistical model checking
approach to an ANIMO model, we introduce variability in the initial quantities of TNF$\alpha$ and EGF
in the model of Figure~\ref{fig:large-model-complete}, letting them be chosen randomly on the interval $\{0, \dots, 100\}$.
we then interrogate the model to know what is the probability that MK2 becomes activated before 40 minutes
have passed. An answer is found whithin XY seconds, and informs us that \dots.
Obtaining an answer to the corresponding exact query (``is it possible to reach a state in which MK2 is active
before 40 minutes?'') would take too long even if using the new model introduced in ANIMO 2.0.

@@

Going beyond the user interface, there are a number of ``pros and cons'' for using ANIMO and \tas\ 
in the biological context. First and foremost, being \tas\ an executable formal language,
a state space can be derived from a \tas\ model. This means that state space-related analyses
such as model checking can be performed on \tas: as explained in the Results section, this can
be done directly in ANIMO, bypassing the need for a deeper understanding of \tas\ as
ANIMO acts as an intermediary towards the powerful model checking tool UPPAAL.

The smaller amount of parameters needed to build an ANIMO model, if compared e.g. to
a differential equation-based approach, makes the modelling process faster, allowing
the biologist to obtain interesting results in less time. However, a less
parameter-intensive modelling approach is also less precise: in our case we abstract
sequences of biochemical reactions into activation or inactivation interactions,
simplifying any relation between two entities in a network as either activating or
inhibiting. While this allows for some simplifications in the networks and the underlying
formal models, it does not allow to represent in full detail more specific concepts
such as reactant concentration or reaction affinity. In case such concepts are of
primary importance, we encourage the interested reader to turn to more precise
modelling approaches, such as those based on differential equations.

All reactants modelled in ANIMO are assumed to be present at the same
(unitary) concentration. While this assumption is not always applicable, it encourages
abstract thought: many biological processes can be represented as networks driven by
\emph{activity-based} interactions. Even if with a limited scope, ANIMO can be applied
also in the analysis of metabolic processes, considering the production of protein $A$ as
the activation of node $A$, with degratation being represented as inhibition. As stated before,
it cannot be expected from such models to be a completely realistic representation of their target biological
processes, but they can still be an useful tool. This can be seen for example in the Drosophila
model in the Results section, where mRNA and protein concentrations were abstracted to
activity-driven processes in the ANIMO model.

\section*{Conclusions}
{\Large [[Possibly further reduce these subsections]]}
% The ultimate aim of research projects is to solve a problem or get
% the answer to a scientific question. In biology, an in-depth understanding of
% the relevant biological system is an important step towards this goal. Successive 
% repetitions of the empirical cycle result in a stepwise increase in understanding,
% until the goal is reached. For complex biological systems, computational modelling is 
% indispensable in this process. The mere act of creating 
% a computational model based on prior knowledge, experimental data and hypotheses 
% assists in gaining more insight in the system. 
% 
% %Furthermore, \emph{in silico} experiments 
% %with the model can be used to make predictions that cannot be made by the human brain. 
% 
% We developed a new modelling approach by proposing a series of abstractions from the detailed 
% molecular mechanisms of biological systems. These abstractions reduce the need for kinetic 
% parameters, while preserving enough expressivity for a useful description of the dynamic 
% behaviour of biological networks. A novel modelling tool, ANIMO, allows 
% effective use of this approach and enables an intuitive construction of formal models.

ANIMO is not the first modelling tool to provide an interface to a
modelling formalism. Such interfaces exist in many other tools (as shown in Tab.~\ref{tab:tool-comparison}). With its
focus on user-friendliness and intuitive modelling, ANIMO's main contribution lies 
in making computational modelling more accessible to experts in biology.
Making use of the visual
interface provided by Cytoscape, network representations subscribe to biological conventions. 
Model parameters are kept to a minimum and can be directly accessed by mouse-clicking on 
nodes and edges. Because of the automatic translation of the network topology and 
user-defined parameters into an underlying formal model, training in the use of formal methods 
is not needed.

\subsection*{Final remarks on the models}
In the Results section, we described the construction of an ANIMO
model of the ciracadian clock in \emph{Drosophila Melanogaster}. This model
captured the dynamics of the regulatory network and led to similar 
conclusions as an ODE model that had been
published previously~\cite{drosophila-ode-model}. This finding supports the use of
the series of modelling abstractions on which ANIMO is based. The biggest
difference between the construction of these models is that the model by~\cite{drosophila-ode-model}
is constructed by writing a system of mathematical equations, together
with an algorithm for simulation. In ANIMO, instead, a number of network
nodes is drawn for the molecules involved. 
These nodes are then linked by directed
interactions that represent cause-and-effect relationships, with a single parameter 
that defines the strength of each
interaction. This is a more intuitive approach to construct a model.
Further contributing to an interactive modelling process
is the compositionality of the model. Each node in the network
can be disabled at any time by the user, or extra nodes can be added,
without having to change any of the existing interactions.

In the Results section we also showed the construction of an executable model
of signalling events downstream of
TNF$\alpha$ and EGF in human colon carcinoma cells. This data set has been used for
previous modelling studies, based on partial least-squares regression and fuzzy 
logic~\cite{pathway-leastsquare,pathway-fuzzy}.
The partial least-squares regression model describes an abstract data-driven model 
that uses statistical correlations
to relate signal transduction events to various cellular decisions. This type of modelling is
very useful in uncovering new and unexpected relations. It is also successful in making
predictions, but gives little direct in the dynamic behaviour of the network. Fuzzy
logic analysis led to a model that gave a better fit to the dynamic network behaviour than
discrete logic (Boolean) models. Inspection of the inputs to the logical gates that were used
to model protein behaviour led to the prediction of novel interactions between proteins,
showing the usefulness of this approach. For most of the proteins, such as JNK1, time was
used as an input parameter. For example, the logical gates ``if TNF$\alpha$ is high
\emph{AND} time is low, then JNK1 is high'' and ``if TNF$\alpha$ is high \emph{AND} time is
high, then JNK1 is low'' were used to
describe the dynamic behaviour of JNK1. Although this leads to a representative
description of the dynamic behaviour of JNK1, peaks in protein activity at early time points were
not reproduced by the fuzzy logic model. Moreover, it gives no insight in the molecular interactions 
that are involved in activation or inhibition.

@@

\subsection*{ANIMO 2.0}
We have presented here how the new and improved ANIMO 2.0 
provides an interactive approach to model analysis and a way of exploiting the power of model checking.
Thanks to the increased performances,
we are able to obtain answers to model checking queries in a matter of seconds.
Coherently with the user friendliness of the tool, we ensure that the biologist can access the features
of model checking without the need to directly deal with \tas\ models.
In this way, ANIMO acts as an intermediary between the biologist and a formal
representation of biological signaling pathways, letting the experts concentrate
on investigating the mechanisms of cellular responses.

\subsection*{Future plans}
In order to enforce the concept of user interaction as a primary focus of the tool, we plan to extend
ANIMO by adding support for parameter sensitivity analysis and automatic parameter sweeps.
We already allow for only one numeric parameter per reaction, but an automatic way of adjusting such
parameters to fit experimental data will allow the user to concentrate more on defining a sensible topology for a pathway.
Such an automatic search could be limited by setting specific bounds on reaction rates based on previous knowledge,
or on a trade-off with performances.
Moreover, inspired by works on automata learning~\cite{test-based-modelling}, we plan to add also the possibility
to automatically derive a network topology based on experimental data and user-defined restrictions based on
previous knowledge.
